3D Stellar Obliquity Distribution of Exoplanetary Systems

### 2022-11-23 src/tex/ms.tex
[showyourwork] overleaf sync

% Abstract with filler text
\begin{abstract}
We build a hierarchical Bayesian framework to infer the 3D stellar obliquity $\psi$ distribution in exoplanetary systems. 
First, we apply our framework to \numistar systems with sky-projected stellar obliquity $\lambda$ and stellar inclinations $i_\star$ measurements. The $\cos{\psi}$ distribution has a spike at 1, suggesting the majority of systems are well aligned, and a concentration at $-0.15$ ($\psi \sim 100\degr$), consistent with the finding of a preponderance of perpendicular planets.
Interestingly, if we assume isotropic priors on $i_\star$ and use no extra data, a similar $\cos{\psi}$ distribution is inferred, suggesting the $cos\psi$ distribution depends mainly on the underlying $\lambda$ distribution.
We then apply the framework to all \numall systems with sky-projected obliquity measurements. Different from the observed-$i_\star$ sample, $\cos{\psi}$ is nearly uniformly distributed between $-0.75$ to $0.75$, corresponding to an isotropic distribution of $\psi$ between $40\degr$ and $140\degr$.
To understand the $\cos{\psi}$-distribution discrepancy between two samples, we look at their $\lambda$ distributions. The measured-$i_\star$ sample has a concentration of $\lambda$ at $110\degr$, whereas the full sample's $\lambda$ is uniformly distributed from $30\degr$ to $180\degr$.
The observed-$i_\star$ sample could be biased because misaligned systems are often found around hot stars; however, hot stars' inclinations are mostly measured by the gravity-darkening method, which is biased toward measuring perpendicular planets.
Lastly, we apply the framework to constrain Hot Jupiters' 3D stellar obliquity distribution and find that in misaligned Hot Jupiter systems, the stellar spin axis is isotropically distributed around the planet's orbital axis.
\end{abstract}

@@

3D stellar obliquity $\psi$ is described as the angle between a planet's orbital axis $\bf{i}_{\rm orb}$ and its host star's spin axis $\bf{i}_{\star}$.
For any given $\bf{i}_{\rm orb}$ axis, if $\bf{i}_{\star}$ is a random vector with uniform distribution on a three-dimensional sphere, the probability distribution $p_{\psi} \sim \sin{\psi}$ and thus $p_{\cos{\psi}} = p_{\psi} \abs{\dv*{\psi}{\cos{\psi}}}$ and is uniformly distributed.
In reality, we only measure the sky-projected stellar obliquity $\lambda$, the angle between the projections of $\bf{i}_{\rm orb}$ and $\bf{i}_{\star}$ onto the plane of the sky. Stellar obliquity $\psi$ can be calculated from $\bf{i}_{\rm orb}$ and $\bf{i}_{\star}$ as the following:
\begin{equation}\label{eqn:psi}
    \cos{\psi} = \sin{i_\star}\sin{i_{\rm orb}}\cos{\lambda} + \cos{i_\star}\cos{i_{\rm orb}},
\end{equation}
where $i_{\rm orb}$ is the angle between the vector $\bf{i}_{\rm orb}$ and the observer's line of sight, and $i_\star$ is the angle between $\bf{i}_{\star}$ and the observer's line of sight. A full derivation of Equation~(\ref{eqn:psi}) can be found in, e.g., \cite{Fabrycky09}.
Since the sky-projected stellar obliquities can only be measured in transiting exoplanet systems, $i_{\rm orb} \approx 90\degr$ and $\cos{\psi} \approx \sin{i_\star}\cos{\lambda}$.

To constrain the 3D stellar obliquity $\psi$, both a measurement of the sky-projected obliquity $\lambda$ and the stellar inclination $i_\star$ are required. The stellar inclination can often be constrained on exoplanetary systems via, e.g., photometric and spectroscopic rotational modulation introduced by starspots, gravity darkening, and asteroseismology. For systems without measured $i_\star$, it is still possible to infer the 3D stellar obliquity $\psi$ from the sky-projected obliquity $\lambda$ assuming the stellar inclination $i_\star$ is isotropically distributed \citep[e.g.,][]{Fabrycky09}. The probability distribution follows $p_{i_\star} \sim \sin{i_{\star}}$, or $p_{\cos{i_\star}} \sim \mathcal{U}$.

\cite{Albrecht21} inferred the 3D stellar obliquities of a sample of stars that have both stellar inclination and sky-projected stellar obliquity constraints. They found while the majority of planets have a small $\psi$, misaligned planets have $\psi$ clustered between 80--125$\degr$. Furthermore, the $\cos{\psi}$ distribution of misaligned systems is not uniformly distributed but clusters near 0, suggesting the stellar spin axis is not isotropically distributed around the planetary orbital axis.

In this work, we build a hierarchical Bayesian framework to evaluate the distribution of $\cos{\psi}$ using the sample in \cite{Albrecht21} and extend the analysis to the full sample of exoplanetary systems that have sky-projected stellar obliquity measurements extracted from the \cite{Albrecht22} review.
In Section~\ref{sec:hbm}, we describe our hierarchical Bayesian framework and the inferred 3D stellar obliquity distributions, and in Section~\ref{sec:interpretations}, we discuss the interpretations of these distributions. Lastly, in Section~\ref{sec:hjs}, we apply the framework to Hot Jupiters and infer their 3D stellar obliquity distribution.

\section{3D Stellar Obliquity Distributions} \label{sec:hbm}
We build a hierarchical Bayesian framework to model the 3D stellar obliquity ($\psi$) distribution of exoplanetary systems.

First, we apply the framework to the sample of \numistar systems that have both sky-projected stellar obliquity $\lambda$ and stellar inclination $i_\star$ constraints.
In Figure~\ref{fig:graph}, we illustrate the probabilistic graphical model. We aim to constrain a set of hyperparameters $\bm{\beta}$ that describe the 3D stellar obliquity $\psi$ distribution. $\bm{\beta}$ is constrained by $N$ individual systems, where each system has $\psi_n$ that is calculated by the sky-projected stellar obliquity $\lambda_n$ and the stellar inclination $i_n$. For simplicity, we approximate $i_{\rm orb}$ to $90\degr$ in the model and use Equation~(\ref{eqn:psi}) to compute $\psi_n$. 
The parameter $\theta_{\star, n}$ contains information on stellar rotation period and stellar radius, if available.
If the rotational modulation method is used to constrain the stellar inclination \citep[e.g.,][]{Masuda20}, $i_{\star,n}$ and $\theta_{\star,n}$ are constrained by observed sky-projected stellar rotational line broadening $\hat{v}\sin{i_{\star,n}}$ and observed $\hat{\theta}_{\star,n}$. Lastly, the sky-projected stellar obliquity $\lambda_n$ is constrained by observed $\hat{\lambda}_n$.

\begin{figure}
    \script{graph.py}
    \includegraphics[width=\linewidth]{figures/graph.pdf}
    \caption{Hierarchical Bayesian framework to infer the stellar obliquity distribution of exoplanetary systems. The stellar obliquity distribution is described by hyperparameters $\vb*{\beta}$ and constrained by $n_{\rm th}$ system's stellar obliquity $\psi_n$. Each $\psi_n$ is calculated by the system's sky-projected stellar obliquity $\lambda_n$ and stellar inclination $i_{\star, n}$. Obs$_{\star,n}$ contains observations that help to constrain $i_{\star,n}$, such as photometric or spectroscopic time series and stellar spectra. $\theta_{\star, n}$ contains other information of the star other than $i_{\star, n}$, such as the stellar radius and stellar rotation period and $\hat{\theta}_{\star,n}$ contains observations that are not already included in Obs$_{\star,n}$.}
    \label{fig:graph}
\end{figure}

For the stellar obliquity distribution, we adopt a two-component mixture Beta distribution with hyperparameters $\bm{\beta} = \{\bm{w},\bm{a},\bm{b}\}$, where each hyperparameter has a dimension of 2. The Beta distribution is flexible to describe both the spike of aligned systems with 3D stellar obliquities close to $0\degr$ and the clustered or broadly distributed systems with large obliquities.
The probability distribution follows 
\begin{align}
    \cos{\psi} &\sim 2\times\Bigl( w_0 {\rm Beta}(a_0, b_0) + w_1 {\rm Beta}(a_1, b_1)\Bigr)-1.
\end{align}
Since a Beta distribution is defined on the interval $[0, 1]$, we extend its support to $[-1,1]$ using a linear transformation (i.e., $2\times Beta-1$).

@@

The likelihood functions are the following:
\begin{align}
    L(\lambda) &\sim \prod_{i=1}^N\mathcal{N}(\hat{\lambda}_n, \sigma_{\hat{\lambda}_n}) \nonumber\\
    L(i_\star, \theta_\star) &\sim \prod_{i=1}^N\mathcal{N}(\hat{v}\sin{i_{\star,n}}, \sigma_{\hat{v}\sin{i_{\star,n}}}).
\end{align}

@@

We apply the model to \numistar systems with both constrained $\lambda$ and $i_\star$ summarized in \cite{Albrecht21}. The inferred $\cos{\psi}$ distribution is shown in the left panel of Figure~\ref{fig:psi_dist}. The distribution presents a pike at 1 as expected, suggesting the majority of exoplanetary systems are well-aligned systems. For misaligned systems, $\cos{\psi}$ presents a cluster near $\cos{\psi} = -0.2$, corresponding to $\psi$ of $100\degr$. The result is consistent with the findings of a preponderance of perpendicular planets in \cite{Albrecht21}.

@@

\subsection{Do \texorpdfstring{$i_\star$}{istar} measurements matter?}
One interesting result is with or without $i_\star$ measurements, the $\cos{\psi}$ distributions are very similar, as shown in the left and middle panels of Figure~\ref{fig:psi_dist}. Here we perform some tests to evaluate the importance of $i_\star$ constraints.

@@

When look at the underlying $\lambda$ distribution, we find a cluster of systems centered at $110\degr$.

>>>>

% Abstract with filler text
\begin{abstract}
Stellar obliquity is a key property that implies the formation and dynamical processes of a planetary system.
We build a hierarchical Bayesian framework to infer the 3D stellar obliquity $\psi$ distribution in exoplanetary systems. 
First, we apply our framework to \numistar systems with sky-projected stellar obliquity $\lambda$ and stellar inclinations $i_\star$ measurements. The $\cos{\psi}$ distribution has a sharp peak at 1, suggesting the majority of systems are well aligned, and is clustered at $-0.15$ ($\psi \sim 100\degr$), consistent with the finding of a preponderance of perpendicular planets.
Interestingly, if we simply assume isotropic priors on $i_\star$ and use no additional data, a similar $\cos{\psi}$ distribution is inferred, suggesting the $cos\psi$ distribution depends mainly on the underlying $\lambda$ distribution.
We then apply the framework to all \numall systems with sky-projected obliquity measurements. Unlike the observed-$i_\star$ sample, the $\cos{\psi}$ is nearly uniformly distributed between $-0.75$ to $0.75$, corresponding to an isotropic distribution of $\psi$ between $40\degr$ and $140\degr$.
To understand the discrepancy, we look at the underlying $\lambda$ distributions of two samples. For misaligned systems, the $\lambda$ distribution of the observed-$i_\star$ sample is clustered at $110\degr$, whereas the $\lambda$ distribution of the full sample is uniform. We find the observed-$i_\star$ sample could be biased because misaligned systems are often found around hot stars; however, hot stars' inclinations are mostly measured by the gravity-darkening method, which is biased toward detecting perpendicular planets.
Lastly, we apply our framework to constrain Hot Jupiters' 3D stellar obliquity distribution and find that in misaligned Hot Jupiter systems, the stellar spin axis is isotropically distributed around the planet's orbital axis.
\end{abstract}

@@

3D stellar obliquity $\psi$ is described as the angle between a planet's orbital axis $\bf{i}_{\rm orb}$ and its host star's spin axis $\bf{i}_{\star}$.
For any given $\bf{i}_{\rm orb}$ axis, if $\bf{i}_{\star}$ is a random vector with uniform distribution on a three-dimensional sphere (i.e., $\bf{i}_{\star}$ is isotropically distributed), the probability distribution $p_{\psi} \sim \sin{\psi}$. Since $p_{\cos{\psi}} = p_{\psi} \abs{\dv*{\psi}{\cos{\psi}}}$, $p_{\cos{\psi}}$ is uniformly distributed.
In reality, we only measure the sky-projected stellar obliquity $\lambda$, the angle between the projections of $\bf{i}_{\rm orb}$ and $\bf{i}_{\star}$ onto the plane of the sky. Stellar obliquity $\psi$ can be computed as:
\begin{equation}\label{eqn:psi}
    \cos{\psi} = \sin{i_\star}\sin{i_{\rm orb}}\cos{\lambda} + \cos{i_\star}\cos{i_{\rm orb}},
\end{equation}
where $i_{\rm orb}$ is the angle between the vector $\bf{i}_{\rm orb}$ and the observer's line of sight, and $i_\star$ is the angle between $\bf{i}_{\star}$ and the observer's line of sight. A full derivation of Equation~(\ref{eqn:psi}) can be found in, e.g., \cite{Fabrycky09}.
Since the sky-projected stellar obliquities can be measured only in transiting exoplanet systems, $i_{\rm orb} \approx 90\degr$ and $\cos{\psi} \approx \sin{i_\star}\cos{\lambda}$.

To constrain the 3D stellar obliquity $\psi$, the measurements of the sky-projected obliquity $\lambda$ and stellar inclination $i_\star$ are required. The stellar inclination can often be constrained via, e.g., photometric and spectroscopic rotational modulation introduced by starspots, gravity darkening, and asteroseismology. For systems without measured $i_\star$, it is still possible to infer the 3D stellar obliquity $\psi$ from the sky-projected obliquity $\lambda$ assuming the stellar inclination $i_\star$ is isotropically distributed \citep[e.g.,][]{Fabrycky09}. The probability distribution of $i_star$ follows $p_{i_\star} \sim \sin{i_{\star}}$, or $p_{\cos{i_\star}} \sim 1$.

\cite{Albrecht21} inferred the 3D stellar obliquities of a sample of stars that have both sky-projected stellar obliquity and stellar inclination constraints. They found while the majority of planets have a small $\psi$, misaligned planets have $\psi$ clustered between 80--125$\degr$. Furthermore, the $\cos{\psi}$ distribution of misaligned systems is not uniformly distributed but clusters near 0, suggesting the stellar spin axis is not isotropically distributed around the planetary orbital axis.

In this work, we build a hierarchical Bayesian framework to evaluate the distribution of $\cos{\psi}$ for the sample in \cite{Albrecht21} and extend the analysis to the full sample of exoplanetary systems that have sky-projected stellar obliquity measurements extracted from the \cite{Albrecht22} review.
In Section~\ref{sec:hbm}, we describe our hierarchical Bayesian framework and the inferred 3D stellar obliquity distributions, and in Section~\ref{sec:interpretations}, we discuss the interpretations of these distributions. In Section~\ref{sec:hjs}, we apply the framework to Hot Jupiters and infer their 3D stellar obliquity distribution.

\section{3D Stellar Obliquity Distributions} \label{sec:hbm}
We build a hierarchical Bayesian framework to model the 3D stellar obliquity ($\psi$) distribution of exoplanetary systems. In Figure~\ref{fig:graph}, we illustrate the probabilistic graphical model. We aim to constrain a set of hyperparameters $\bm{\beta}$ that describe the 3D stellar obliquity $\psi$ distribution. $\bm{\beta}$ is constrained by $N$ individual systems, where each system has $\psi_n$ that is calculated by the sky-projected stellar obliquity $\lambda_n$ and the stellar inclination $i_n$. For simplicity, we approximate $i_{\rm orb}$ to $90\degr$ and use the simplified version of Equation~(\ref{eqn:psi}) to compute $\psi_n$.
The parameter $\theta_{\star, n}$ contains stellar parameters other than $i_\star$, e.g., stellar rotation period and stellar radius, if these information are available.
Obs$_{\star,n}$ describes observations that help to constrain $i_{\star,n}$, such as photometric or spectroscopic time series and the stellar spectra. If the rotational modulation method is used to constrain the stellar inclination \citep[e.g.,][]{Masuda20}, $i_{\star,n}$ and $\theta_{\star,n}$ are constrained by observed sky-projected stellar rotational line broadening $\hat{v}\sin{i_{\star,n}}$ and observed $\hat{\theta}_{\star,n}$. Lastly, the sky-projected stellar obliquity $\lambda_n$ is constrained by observed $\hat{\lambda}_n$.

\begin{figure}
    \script{graph.py}
    \includegraphics[width=\linewidth]{figures/graph.pdf}
    \caption{Hierarchical Bayesian framework to infer the stellar obliquity distribution of exoplanetary systems. The stellar obliquity distribution is described by hyperparameters $\vb*{\beta}$ and constrained by $n_{\rm th}$ system's stellar obliquity $\psi_n$. Each $\psi_n$ is calculated by the system's sky-projected stellar obliquity $\lambda_n$ and stellar inclination $i_{\star, n}$. Obs$_{\star,n}$ contains observations that help to constrain $i_{\star,n}$, such as photometric or spectroscopic time series and the stellar spectra. $\theta_{\star, n}$ contains other information of the star other than $i_{\star, n}$, such as the stellar radius and stellar rotation period and $\hat{\theta}_{\star,n}$ contains observations that are not already included in Obs$_{\star,n}$.}
    \label{fig:graph}
\end{figure}

For the stellar obliquity distribution, we adopt a two-component mixture Beta distribution with hyperparameters $\bm{\beta} = \{\bm{w},\bm{a},\bm{b}\}$, where each hyperparameter has a dimension of 2. The Beta distribution is flexible to describe both the spike of aligned systems with 3D stellar obliquities close to $0\degr$ and the clustered or broadly distributed systems of misaligned systems.
The probability distribution follows 
\begin{align}
    \cos{\psi} &\sim 2\times\Bigl( w_0 {\rm Beta}(a_0, b_0) + w_1 {\rm Beta}(a_1, b_1)\Bigr)-1.
\end{align}
Since a Beta distribution is defined on the interval $[0, 1]$, we extend its support to $[-1,1]$ using a linear transformation (i.e., $2\times Beta-1$).

@@

The likelihood functions are the following:
\begin{align}
    L(\lambda) &\sim \prod_{i=1}^N\mathcal{N}(\hat{\lambda}_n, \sigma_{\hat{\lambda}_n}) \nonumber\\
    L(i_\star, \theta_\star) &\sim \prod_{i=1}^N\mathcal{N}({\rm Obs}_{\star,n}, \sigma_{{\rm Obs}_{\star,n}}).
\end{align}

@@

First, we apply the framework to the sample of \numistar systems that have both sky-projected stellar obliquity $\lambda$ and stellar inclination $i_\star$ constraints summarized in \cite{Albrecht21}. The inferred $\cos{\psi}$ distribution is shown in the left panel of Figure~\ref{fig:psi_dist}. The distribution presents a pike at 1 as expected, suggesting the majority of exoplanetary systems are well-aligned systems. For misaligned systems, $\cos{\psi}$ presents a cluster near $\cos{\psi} = -0.2$, corresponding to $\psi$ of $100\degr$. The result is consistent with the findings of a preponderance of perpendicular planets in \cite{Albrecht21}.

@@

\subsection{Do \texorpdfstring{$i_\star$}{istar} measurements matter?}

There is no doubt that $i_\star$ measurement improves the constraint on $\psi$ for an individual system. However, do $i_\star$ measurements in individual systems matter when we look at the population $\psi$ distribution? Particularly, the left and middle panels in Figure~\ref{fig:psi_dist} show that with or without $i_\star$ measurements, the $\cos{\psi}$ distributions are very similar. Here we perform some tests to evaluate the importance of $i_\star$ constraints.

We begin with writing down some designed $\lambda$ distribution and assume an isotropic stellar inclination distribution. We then randomly draw 100 samples from these two distributions and assume these numbers are \emph{true} values of $\lambda$ and $i_\star$. We choose 100 so that the sample is large enough to represent the distributions, although, in reality, the number of misaligned systems with $i_\star$ measurements is much smaller ($\sim 20$). Now each $\lambda$ and $i_\star$ is added to some Gaussian noises to simulate \emph{observed} $\lambda$ and $i_\star$. We use uncertainties $\sigma_{\lambda} = 8\degr$ and $\sigma_{i_\star} = 10\degr$ that represent typical observation uncertainties of misaligned systems. Using the observed $\lambda$ and $i_\star$ and their uncertainties, we infer the $\cos{\psi}$ distribution of the sample with or without the $i_\star$ observations. We test three representative distributions of $\cos{\psi}$: (1) $\cos{\psi}$ uniformly distributed from $[-1,1]$, where we have the input $\lambda$ uniformly distributed over $[0, \pi]$, (2) $\cos{\psi}$ clustered at $-0.4$, where we have the input $\lambda$ normally distributed as $\mathcal{N}(120\degr,20\degr)$, and lastly (3) $\cos{\psi}$ clustered at $0$, where we have the input $\lambda$ normally distributed as $\mathcal{N}(90\degr,20\degr)$. In all three distributions, the $\cos{\psi}$ distribution inferred with or without $i_\star$ likelihoods are similar and consistent with the injected $\cos{\psi}$ distributions.

Our experiment is under the assumption that misaligned systems still have isotropic stellar inclinations, which seems to be true.

@@

When looking at the underlying $\lambda$ distribution, we find a cluster of systems centered at $110\degr$.
